Local Tuning in Peano Curves-based Global Optimization Scheme

### 2016-09-25 2016/YSC/ysc_paper.tex
Update YSC paper

The global optimization problem can be formulated as follows: to find the global minimum of an
\(N\)-dimensional function \(\varphi(y)\) in a hyperinterval
\(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\): \(\varphi(y^*)=\min\{\varphi(y):y\in D\}\).
In order to obtain an estimate of the global minimum from a finite number of function value computations,
 \(\varphi(y)\) is required to satisfy Lipschitz condition.
\begin{displaymath}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty
\end{displaymath}
\par
The use of the evolvents \(y(x)\) i.e. the curves filling the space are a classic
dimension-reduction scheme for global optimization algorithms \cite{strOptBook}.
\begin{displaymath}
\label{cube}
\lbrace y\in R^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant N\rbrace=\{y(x):0\leqslant x\leqslant 1\}
\end{displaymath}
\par
Such a mapping allows the reduction of a problem stated in a multidimensional space
to solving a one-dimensional problem at the expense of degrading its properties.
In particular, the one-dimensional function \(\varphi(y(x))\) is not a Lipschitzian but a Hölderian function:
\begin{displaymath}
\label{holder}
|\varphi(y(x_1))-\varphi(y(x_2))|\leqslant H{|x_1-x_2|}^{\frac{1}{N}},x_1,x_2\in[0,1]
\end{displaymath}
where the Hölder constant \(H\) is related to the Lipschitz constant \(L\) by the relation
\begin{displaymath}
H=4Ld\sqrt{N},d=\max\{b_i-a_i:1\leqslant i\leqslant N\}
\end{displaymath}
\par
Therefore, not limiting the generality, one can consider the minimization of the
one-dimensional function \(f(x)=\varphi(y(x))\), with \(x\in[0;1]\) satisfying Hölder condition.
The issues of numerically building the mapping like a Peano curve and the corresponding
theory have been considered in detail in \cite{strOptBook}. Here we would note that an evolvent
built numerically is an approximation to the theoretical Peano curve with a precision
of the order \(2^{-m}\) where \(m\) is the building parameter of the evolvent.

@@

The algorithm considered for solving the stated problem implies building
a sequence of points \(x_k\), in which the values of the minimized function \(z_k = f(x_k)\)
are computed. Let us call the process of computating the function value
(including building an image \(y_k=y(x_k)\)) a trial, and the pair \((x_k,z_k)\) ---
the result of the trial. A set of the pairs \(\{(x_k,z_k)\}, 1\leqslant k\leqslant n\)
makes up the search information accumulated by the method after executing \(n\) steps.
\par
At the first iteration of the method, a trial is executed at an arbitrary internal
point \(x_1\) within the interval \([0,1]\). Let us assume \(n \geqslant 1\) iterations
of the method to be completed, during the course of which the iterations in \(k = k(n)\)
points \(x_i, 1\leqslant i\leqslant k\) have been performed. Then, the point \(x^{k+1}\)
of the search trial for the next \((k+1)\)th iteration is determined in accordance with the rules:
\par
Step 1. Renumber the points in the set \(X_k=\{x^1,\dotsc,x^k\}\cup\{0\}\cup\{1\}\),
which includes the boundary points of the interval \([0,1]\) as well as the points of
preceding trials, by the lower indices in order of increasing coordinate values  i.e.
\begin{displaymath}
0=x_0<x_1<\dotsc<x_{k+1}=1
\end{displaymath}
\par
Step 2. Assuming \(z_i=f(x_i),1\leqslant i\leqslant k\), compute the quantities
\begin{equation}
\label{step2}
\mu=\max_{1\leqslant i\leqslant k}\dfrac{|z_i-z_{i-1}|}{\Delta_i},
\begin{matrix}
    M =
    \left\{
    \begin{matrix}
    r\mu,\mu>0 \\
    1,\mu=0
    \end{matrix} \right.
    \end{matrix}
\end{equation}
where \(r>1\) is a predefined parameter for the method, and \(\Delta_i=(x_i-x_{i-1})^\frac{1}{N}\).
\par
Step 3. For each interval \((x_{i-1},x_i),1\leqslant i\leqslant k+1\), compute the
characteristics according to the formulae
\begin{equation}
\label{step3_1}
R(1)=2\Delta_1-4\dfrac{z_1}{M},R(k+1)=2\Delta_{k+1}-4\dfrac{z_k}{M}
\end{equation}
\begin{equation}
\label{step3_2}
R(i)=\Delta_i+\dfrac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\dfrac{z_i+z_{i-1}}{M},1<i<k+1
\end{equation}
\par
Step 4. Select the interval \((x_{t-1}, x_t)\) such that
\begin{equation}
\label{step4}
t=\argmax_{1\leqslant i \leqslant k+1}R(i),
\end{equation}
i.e., the interval with the maximal characteristic.
\par
Step 5. Execute a new trial at point \(x_{k+1}\) computed according to the formulae
\begin{displaymath}
x_{k+1}=\dfrac{x_{t}+x_{t-1}}{2},t=1,t=k+1
\end{displaymath}
\begin{equation}
\label{step5}
x_{k+1}=\dfrac{x_{t}+x_{t-1}}{2}-\sign(z_{t}-z_{t-1})\dfrac{1}{2r}\left[\dfrac{|z_{t}-z_{t-1}|}{\mu}\right]^N,1<t<k+1
\end{equation}
\par
The algorithm is terminated if the condition \(\Delta_{t}\leqslant \varepsilon\) is fulfilled;
here \(\varepsilon>0\) is the predefined precision. As an estimate of the global optimum solution of the problem the values
\begin{equation}
f_k^*=\min_{1\leqslant i \leqslant k}f(x_i), x_k^*=\argmin_{1\leqslant i \leqslant k}f(x_i)
\end{equation}
are selected. The theoretical substantiation of this method is presented in \cite{strOptBook}.

%------------------------------------------------------------------------------
\section{Local Adaptive Estimate of Hölder Constant}
%\label{sect:algorithm}
As it is seen from the scheme of the algorithm, regardless of the local properties
of the optimized one-dimensional function, the same estimate of the Hölder
constant (\ref{step2}) is used to compute the characteristic of all intervals (\ref{step3_1}), (\ref{step3_2}).
In \cite{sergLocalTuning}, it has been proposed to use different values of \(M\) for each interval,
and also the efficiency of this approach has been shown in the case of optimizing
one-dimensional functions that satisfy the Lipschitz condition. In \cite{nestedLocal},
the application of the adaptive estimates of Lipschitz constants in the multidimensional
nested optimization scheme has been considered. For each interval, the local estimate of the
constant is an additive convolution of the ``global'' and ``local'' constants
(\(\gamma\) and \(\lambda\), correspondingly):
\begin{displaymath}
  \begin{array}{lr}
    \lambda_i=\max\{H_{i-1},H_i,H_{i+1}\} \\
    H_i=\frac{|z_i-z_{i-1}|}{\Delta_i} \\
    H^k=\max\{H_i:i=2,\dots ,k\} \\
    \gamma_i=H^k\frac{\Delta_i}{\Delta^{max}} \\
    \Delta^{max}=\max\{\Delta_{i}:i=2,\dots ,k\}
  \end{array}
\end{displaymath}
\begin{equation}
\label{additiveConv}
M_i=r\cdot \max\{H_i, \frac{1}{2}(\lambda_i+\gamma_i),\xi\}
\end{equation}
A small \(\xi\) is chosen to preventthe function from being constant over the search interval.
This variant of convolution does not depend on the parameter \(r\), however, the adaptive convolution:
\begin{equation}
\label{additiveAdaptiveConv}
M_i=r\cdot \max\{H_i, \frac{\lambda_i}{r}+\frac{r-1}{r}\gamma_i,\xi\}
\end{equation}
has been considered in \cite{sergLocalTuning} as well.
If it is known a priori that the optimized function has a complex shape with multiple
local minima, then the initial value of \(r\) is specified to be large enough to
result in the dominance of the ``global'' component \(\gamma\) in the adaptive convolution.

>>>>

The global optimization problem can be formulated as follows: to find the global minimum of an
\(N\)-dimensional function \(\varphi(y)\) in a hyperinterval
\(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\):
\begin{displaymath}
  \varphi(y^*)=\min\{\varphi(y):y\in D\}.
\end{displaymath}
In order to obtain an estimate of the global minimum from a finite number of function value computations,
 \(\varphi(y)\) is required to satisfy Lipschitz condition.
\begin{displaymath}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty
\end{displaymath}
\par
The use of the evolvents \(y(x)\) i.e. the curves filling the space are a classic
dimension-reduction scheme for global optimization algorithms \cite{evolvents2013}.
\begin{displaymath}
\label{cube}
\lbrace y\in R^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant N\rbrace=\{y(x):0\leqslant x\leqslant 1\}
\end{displaymath}
\par
Such a mapping allows the reduction of a problem stated in a multidimensional space
to solving a one-dimensional problem at the expense of worsening its properties.
In particular, the one-dimensional function \(\varphi(y(x))\) is not a Lipschitzian but a Hölderian function:
\begin{displaymath}
\label{holder}
|\varphi(y(x_1))-\varphi(y(x_2))|\leqslant H{|x_1-x_2|}^{\frac{1}{N}},x_1,x_2\in[0,1]
\end{displaymath}
where the Hölder constant \(H\) is related to the Lipschitz constant \(L\) by the relation
\begin{displaymath}
H=4Ld\sqrt{N},d=\max\{b_i-a_i:1\leqslant i\leqslant N\}
\end{displaymath}
\par
Therefore, not limiting the generality, one can consider the minimization of the
one-dimensional function \(f(x)=\varphi(y(x))\), with \(x\in[0,1]\) satisfying Hölder condition.
The issues of numerically building the mapping like a Peano curve and the corresponding
theory have been considered in detail in \cite{evolvents2013}. Here we would note that an evolvent
built numerically is an approximation to the theoretical Peano curve with a precision
of the order \(2^{-m}\) where \(m\) is the building parameter of the evolvent.

@@

The algorithm considered for solving the stated problem implies generating
a sequence of points \(x_k\), in which the values of the minimized function \(z_k = f(x_k)\)
are computed. Let us call the process of computating the function value
(including calulating an image \(y^k=y(x^k)\)) a trial, and the pair \((x^k,z^k)\) ---
the result of the trial. A set of the pairs \(\{(x^k,z^k)\}, 1\leqslant k\leqslant n\)
makes up the search information accumulated by the method after executing \(n\) steps.
\par
At the first iteration of the method, a trial is executed at an arbitrary internal
point \(x^1\) within the interval \([0,1]\). Let us assume \(n \geqslant 1\) iterations
of the method to be completed, during the course of which the iterations in \(k = k(n)\)
points \(x_i, 1\leqslant i\leqslant k\) have been performed. Then, the point \(x^{k+1}\)
of the search trial for the next \((k+1)\)th iteration is determined in accordance with the rules:
\par
Step 1. Renumber the points in the set \(X_k=\{x^1,\dotsc,x^k\}\cup\{0\}\cup\{1\}\),
which includes the boundary points of the interval \([0,1]\) as well as the points of
preceding trials, by the lower indices in order of increasing coordinate values  i.e.
\begin{displaymath}
0=x_0<x_1<\dotsc<x_{k+1}=1
\end{displaymath}
\par
Step 2. Assuming \(z_i=f(x_i),1\leqslant i\leqslant k\), compute the values
\begin{equation}
\label{step2}
\mu=\max_{1\leqslant i\leqslant k}\dfrac{|z_i-z_{i-1}|}{\Delta_i},
\begin{matrix}
    M =
    \left\{
    \begin{matrix}
    r\mu,\mu>0 \\
    1,\mu=0
    \end{matrix} \right.
    \end{matrix}
\end{equation}
where \(r>1\) is a predefined parameter for the method, and \(\Delta_i=(x_i-x_{i-1})^\frac{1}{N}\).
\par
Step 3. For each interval \((x_{i-1},x_i),1\leqslant i\leqslant k+1\), compute the
characteristics according to the formulae
\begin{equation}
\label{step3_1}
R(1)=2\Delta_1-4\dfrac{z_1}{M},R(k+1)=2\Delta_{k+1}-4\dfrac{z_k}{M},
\end{equation}
\begin{equation}
\label{step3_2}
R(i)=\Delta_i+\dfrac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\dfrac{z_i+z_{i-1}}{M},1<i<k+1.
\end{equation}
\par
Step 4. Select the interval \((x_{t-1}, x_t)\) such that
\begin{equation}
\label{step4}
t=\argmax_{1\leqslant i \leqslant k+1}R(i),
\end{equation}
i.e., the interval with the maximal characteristic.
\par
Step 5. Execute a new trial at point \(x_{k+1}\) computed according to the formulae
\begin{displaymath}
x_{k+1}=\dfrac{x_{t}+x_{t-1}}{2},t=1,t=k+1,
\end{displaymath}
\begin{equation}
\label{step5}
x_{k+1}=\dfrac{x_{t}+x_{t-1}}{2}-\sign(z_{t}-z_{t-1})\dfrac{1}{2r}\left[\dfrac{|z_{t}-z_{t-1}|}{\mu}\right]^N,1<t<k+1.
\end{equation}
\par
The algorithm is terminated if the condition \(\Delta_{t}\leqslant \varepsilon\) is fulfilled;
here \(\varepsilon>0\) is the predefined accuracy. As an estimate of the global optimum solution of the problem the values
\begin{equation}
f_k^*=\min_{1\leqslant i \leqslant k}f(x_i), x_k^*=\argmin_{1\leqslant i \leqslant k}f(x_i)
\end{equation}
are selected. The theoretical substantiation of this method is presented in \cite{strOptBook}, chapter 8.

%------------------------------------------------------------------------------
\section{Local Adaptive Estimate of Hölder Constant}
%\label{sect:algorithm}
As it is seen from the scheme of the algorithm, regardless of the local properties
of the optimized one-dimensional function, the same estimate of the Hölder
constant (\ref{step2}) is used to compute the characteristic of all intervals (\ref{step3_1}), (\ref{step3_2}).
In \cite{sergLocalTuningFirst}, it has been proposed to use different values of \(M\) (\(M\) is the same as in (\ref{step2}), (\ref{step3_1})) for each interval,
and also the efficiency of this approach has been shown in the case of optimizing
one-dimensional functions that satisfy the Lipschitz condition. In \cite{nestedLocal},
the application of the adaptive estimates of Lipschitz constants in the multidimensional
nested optimization scheme has been considered. For each interval, the local estimate of the
constant is an additive convolution of the ``global'' and ``local'' constants
(\(\gamma\) and \(\lambda\), correspondingly):
\begin{displaymath}
  \begin{array}{lr}
    \lambda_i=\max\{H_{i-1},H_i,H_{i+1}\} \\
    H_i=\frac{|z_i-z_{i-1}|}{\Delta_i} \\
    H^k=\max\{H_i:i=2,\dots ,k\} \\
    \gamma_i=H^k\frac{\Delta_i}{\Delta^{max}} \\
    \Delta^{max}=\max\{\Delta_{i}:i=2,\dots ,k\}
  \end{array}
\end{displaymath}
\begin{equation}
\label{additiveConv}
M_i=r\cdot \max\{H_i, \frac{1}{2}(\lambda_i+\gamma_i),\xi\}
\end{equation}
A small \(\xi\) is chosen to prevent the function from being constant over the search interval.
This variant of convolution does not depend on the parameter \(r\), however, the adaptive convolution:
\begin{equation}
\label{additiveAdaptiveConv}
M_i=r\cdot \max\{H_i, \frac{\lambda_i}{r}+\frac{r-1}{r}\gamma_i,\xi\}
\end{equation}
has been introduced in \cite{gergel2010} and detailed considered (in case of solving one-dimentional Lipschitzian problems) in \cite{sergLocalTuning} as well.
If it is known a priori that the optimized function has a complex shape with multiple
local minima, then the initial value of \(r\) is specified to be large enough to
result in the dominance of the ``global'' component \(\gamma\) in the adaptive convolution.
